---
output:
  html_document:
    keep_md: yes
---
<!--
    https://rstudio-pubs-static.s3.amazonaws.com/35364_334944a02fcf4792b232aacbfee2f6e6.html
    https://rpubs.com/roozbehdavari/33121
-->
# Assessing the ToothGrowth Dataset with R
#### Author: Danilo Mutti

```{r, echo = FALSE}
knitr::opts_chunk$set(fig.path = "course_project_b_figure/")
set.seed(1)
library(ggplot2, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
```

The [ToothGrowth dataset](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html) (TG) contains data about the effect of vitamin C on tooth growth in Guinea pigs. In this study, the response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

The dataset contains a data frame with 60 observations on 3 variables.

|column index|column name| type      | description |
|------------|-----------|-----------|-------------|
| [,1]       | len	     | numeric	 |Tooth length |
| [,2]       | supp	     | factor    | Supplement type (VC or OJ).|
| [,3]       | dose      | numeric   | Dose in milligrams.|

We start by loading the dataset, followed by a basic exploration of how the data is structured.

```{r}
data(ToothGrowth)
str(ToothGrowth)
summary(ToothGrowth)
```

The following figure *seems to indicate* that orange juice (OJ) is more effective for lower/intermediate dose levels. When dose levels are higher, the effectiveness between OJ and Vitamin C (VC) *seems to be similar*, with OJ presenting more consistent results.

```{r}
coplot(len ~ dose | supp, data = ToothGrowth, panel = panel.smooth,
       xlab = "ToothGrowth data: length vs dose, given type of supplement")
```


## Data Processing

In this dataset, rows 1 to 10 correspond to the observations from Guinea pigs 1 to 10. The eleventh row contains another observation of Guinea pig number 1, and so on. In order to tidy this dataset, we are going to fit it into 6 columns and 10 rows. Each row corresponds to one specific Guinea pig, and each column represents a supplement (VC or OJ) and its associated dose (0.5, 1.0 or 2.0).

```{r}
VC_05 <- ToothGrowth$len[1:10]
VC_10 <- ToothGrowth$len[11:20]
VC_20 <- ToothGrowth$len[21:30]
OJ_05 <- ToothGrowth$len[31:40]
OJ_10 <- ToothGrowth$len[41:50]
OJ_20 <- ToothGrowth$len[51:60]
print(tidy <- data.frame(VC_05, VC_10, VC_20, OJ_05, OJ_10, OJ_20))
```

## Data Analysis

## Appendix

This section contains the environment in which this report was produced.

```{r}
sessionInfo()
```