---
output:
  pdf_document: default
  html_document:
    keep_md: yes
---
<!--
    https://rstudio-pubs-static.s3.amazonaws.com/35364_334944a02fcf4792b232aacbfee2f6e6.html
    https://rpubs.com/roozbehdavari/33121
    http://www.r-bloggers.com/two-sample-students-t-test-1/
-->
# Assessing the ToothGrowth Dataset with R
#### Author: Danilo Mutti

```{r, echo = FALSE}
knitr::opts_chunk$set(fig.path = "course_project_b_figure/", fig.width = 5, fig.height = 3, dpi = 144)
set.seed(1)
library(ggplot2, warn.conflicts = FALSE)
library(reshape2, warn.conflicts = FALSE)
```

The [ToothGrowth dataset](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html) (TG) contains data about the effect of vitamin C on tooth growth in Guinea pigs. In this study, the response is the length of odontoblasts (teeth) in each of 60 distinct Guinea pigs, 10 for each combination of dose level of Vitamin C (0.5, 1, and 2 mg) and delivery method (orange juice or ascorbic acid).

The dataset contains a data frame with 60 observations on 3 variables.

| Column Index | Column Name | Column Type | Description                 |
|:------------:|:-----------:|:-----------:|-----------------------------|
| [,1]         | len	     | numeric	   | Tooth length                |
| [,2]         | supp	     | factor      | Supplement type (VC or OJ). |
| [,3]         | dose        | numeric     | Dose in milligrams.         |

We start by loading the dataset, followed by a basic exploration of how the data is structured.

```{r}
data(ToothGrowth)
str(ToothGrowth)
```

The following figure *seems to indicate* that orange juice (OJ) is more effective for lower/intermediate dose levels. When dose levels are higher, the effectiveness between OJ and Vitamin C (VC) *seems to be similar*, with OJ presenting more consistent results.

```{r}
g <- ggplot(data = ToothGrowth, aes(x = interaction(supp, dose), y = len, fill = supp))
g <- g + geom_boxplot(outlier.colour = "red") + scale_fill_brewer(palette = "Pastel2")
g <- g + labs(title="ToothGrowth by dose and supplement type",
              x="Supplement type and dose", y="Tooth length"); g
```

## Data Processing

In this dataset, rows 1 to 10 correspond to the observations from Guinea pigs 1 to 10. The eleventh row contains another observation of Guinea pig number 1, and so on. In order to tidy this dataset, we are going to fit it into 6 columns and 10 rows. Each row corresponds to one specific Guinea pig, and each column represents a supplement (VC or OJ) and its associated dose (0.5, 1.0 or 2.0).

```{r}
ToothGrowth$guinea.pig <- rep(c(1:10), 6)
print(tidy <- dcast(ToothGrowth, guinea.pig ~ supp + dose, value.var = "len"))
```

## Data Analysis

In this section, we are going to perform several tests to compare tooth growth of Guinea pigs by supplement (OJ, VC) and dose (0.5, 1.0, and 2.0 milligrams).
For every comparation, we will consider the following hypotheses:

* H_0: Both supplements, for a given dosage, have the same effectiveness
* H_a: OJ is more effective than VC, for a given dosage.

The `t.test` function requires two other parameters: `var.equal` and `paired`. The first parameter indicates whether to treat the two variances as being equal. If TRUE then the pooled variance is used to estimate the variance otherwise the Welch (or Satterthwaite) approximation to the degrees of freedom is used. The dataset provides no information on this subject, hence we consider the safer choice and take the equality as FALSE. The second parameter is set as FALSE, due to the description of the dataset, which states that there were 60 distinct Guinea pigs in the experiment.

### 0.5 milligrams - OJ versus VC

```{r}
t.test(tidy$OJ_0.5, tidy$VC_0.5, paired = FALSE, var.equal = FALSE)
```

The `t.test` result **rejects**, with a 95% confidence interval, the null hypothesis H_0 of equality of effectiveness, i.e., OJ is more effective than VC when the dosage is 0.5mg.

### 1.0 milligrams - OJ versus VC

```{r}
t.test(tidy$OJ_1, tidy$VC_1, paired = FALSE, var.equal = FALSE)
```

Once again, the `t.test` result **rejects**, with a 95% confidence interval, the null hypothesis H_0 of equality of effectiveness. Orange juice is more effective than Vitamin C when the dosage is 1.0mg.


## Appendix

This section contains the environment in which this report was produced.

```{r}
sessionInfo()
```