---
output:
  html_document:
    keep_md: yes
---
1. Suppose that diastolic blood pressures (DBPs) for men aged 35-44 are normally distributed with a mean of 80 (mm Hg) and a standard deviation of 10. About what is the probability that a random 35-44 year old has a DBP less than 70?

```{r}
pnorm(70, mean = 80, sd = 10, lower.tail = TRUE)
```

2. Brain volume for adult women is normally distributed with a mean of about 1,100 cc for women with a standard deviation of 75 cc. What brain volume represents the 95th percentile?
```{r}
qnorm(.95, mean = 1100, sd = 75)
```

3. Refer to the previous question. Brain volume for adult women is about 1,100 cc for women with a standard deviation of 75 cc. Consider the sample mean of 100 random adult women from this population. What is the 95th percentile of the distribution of that sample mean?

```{r}
1100 + (75/sqrt(100) * 1.645)
```

4. You flip a fair coin 5 times, about what's the probability of getting 4 or 5 heads?

```{r}
pbinom(3,size=5,prob=.5,lower.tail = FALSE)
```

5. The respiratory disturbance index (RDI), a measure of sleep disturbance, for a specific population has a mean of 15 (sleep events per hour) and a standard deviation of 10. They are not normally distributed. Give your best estimate of the probability that a sample mean RDI of 100 people is between 14 and 16 events per hour?

```{r}
mean <- 15
sd <- 10
value1 <- (14-mean) / (sd/sqrt(100))
p14 <- pnorm(value1)
value2 <- (16-mean) / (sd/sqrt(100))
p16 <- pnorm(value2)
p16-p14
```

6. Consider a standard uniform density. The mean for this density is .5 and the variance is 1 / 12. You sample 1,000 observations from this distribution and take the sample mean, what value would you expect it to be near?

```
mean = 1/2 * (min + max)
1/2 = 1/2 * (min + max)
1 = min + max
```

```
var = 1/2 * ((max - min) ^ 2)
1/12 = 1/12 * ((max - min) ^ 2)
1 = (max - min) ^ 2
````

[Source](http://en.wikipedia.org/wiki/Uniform_distribution_(continuous))

```{r}
min <- .5
max <- 1.5

mean(cumsum(runif(1000, min = min, max = max))/1000)
```

7. The number of people showing up at a bus stop is assumed to be Poisson with a mean of 5 people per hour. You watch the bus stop for 3 hours. About what's the probability of viewing 10 or fewer people?

```{r}
ppois(10, lambda = 3 * 5)
```